TRIK Studio: Technical Introduction

### 2017-02-09 2017-trikStudioEnglish/trikStudioEnglish.tex
translated 2D model

The data-flow interpreter works in two steps. First, the interpreted diagram is being validated, implementation objects for diagram blocks are created (they are used to translate the diagram into a sequence of robot commands similar to as it was discussed for the control-flow programs interpreter). Implementation objects are connected with each other via Qt signals and slots in the same order as corresponding blocks are connected on a diagram (<<Publisher-subscriber>> design pattern is used here). On the second step the interpreter executes each block that does not have any incoming data flows, and then each such block in its turn activates blocks that it has outgoing data flows to when the data tokens are ready. Blocks execution is performed in a pseudo-parallel way over a centralized message queue (see Fig.~\ref{image:dsPseudoparallelism}). This implementation detail distinguishes this language from all other industrial solutions (e.g. in Microsoft Robotics Developer Studio diagrams are deployed as a set of independent web services~\cite{jackson2007microsoft). We choose this approach because this programming environment targets low-performance controllers of educational robots, that don't embrace true parallel execution of multiple programs at once. Nevertheless, the language has a process fork block that is useful while implementing complex control models mentioned above. This block should be used as a low-level tool for execution control.

@@

Особая часть среды TRIK Studio, отделимая от всей остальной системы --- подсистема двумерного имитационного моделирования робота (двумерный симулятор). Окно симулятора встроено в интерфейс TRIK Studio, но может и использоваться отдельно. Основная часть симулятора --- редактор модели мира. В специальном меню можно выбрать инструмент рисования (подобно тому, как это происходит в большинстве графических редакторов). Инструменты рисования делятся на стены (твердые предметы, на которые реагируют ультразвуковые и инфракрасные датчики роботов и сквозь которые робот не может проехать) и цветные маркеры на полу (элементы, на которые реагируют датчики цвета, освещенности и эмуляторы систем видео-зрения) --- прямые линии и кубические кривые безье, прямоугольники, эллипсы и стилус (произвольная растровая фигура, рисующаяся мышью как карандашом). Для каждого маркера на полу может быть задана его толщина, цвет и заливка. Модель робота всегда присутствует на сцене редактора мира, на ней произвольным образом можно размещать виртуальные датчики. Робот представляет собой двухколесную тележку с дифференциальным приводом. Для удобства регионы сканирования датчиков расстояния подсвечиваются. На отдельной панели для каждой поддерживаемой модели робота (NXT, EV3 или TRIK) присутствует эмулятор панели контроллера: фронтальное изображение его лицевой панели с кнопками, нажатие мышью на которые эмулирует нажатие на кнопку реального контроллера, цветными светодиодами, меняющими свой цвет соответственно тому, как того требует написанная пользователем программа, и эмулятором дисплея, изображение на котором можно менять блоками из группы <<Рисование>>.

Важной особенностью симулятора является то, что режим исполнения программы неотличим от режима создания модели мира. Модель мира может редактироваться даже в момент исполнения программы, на любое добавление пользователем стенок и цветных элементов во время исполнения программы датчики начнут реагировать немедленно; положение и направление робота и его датчиков в пространстве может быть изменено в любой момент, что соответствует в реальном мире физическому воздействию на корпус робота.

Симулятор построен на основе архитектуры Model-View-Controller. Модель содержит сериализуемое описание мира и робота, уведомляет о любом изменении любого свойства любого предмета. На события модели подписывается представление, все изменения в модели автоматически отображаются на сцене и панелях симулятора. Пользовательские действия в представлении выполняются через контролирующую компоненту, которая кладет очередное действие на вершину специального стека для возможности его отмены и повтора.

Важной частью модели является ее <<физика>>. В симуляторе реализовано две физические модели поведения робота: идеальная и реалистичная. В идеальной физике игнорируются силы трения о пол и стены, импульсы моторов, при неизменных скоростях моторов робот будет двигаться равномерно (без ускорения и замедления). Любое, даже самое легкое столкновение со стеной в такой модели остановит робота. В реалистичной модели ведется полный учет сил тяги и трения робота о пол и стены, при столкновении со стеной робот поведет себя подобно тому, как это происходит в реальности. Переключение между физическими моделями происходит при выставлении соответствующего флага на панели настроек симуляции и может произойти даже во время исполнения программы. Имеется возможность <<зашумления>> показаний датчиков и импульсов на моторах гауссовым шумом.

Время в двумерном симуляторе не соответствует процессорному: в нем существует централизованная временная прямая, темп хода которой может меняться на специальной панели (что не меняет поведения робота). Та же самая временная прямая используется в интерпретаторах визуальных языков TRIK Studio для соответствия блоков работы со временем модельному, а не процессорному времени.

>>>>

The data-flow interpreter works in two steps. First, the interpreted diagram is being validated, implementation objects for diagram blocks are created (they are used to translate the diagram into a sequence of robot commands similar to as it was discussed for the control-flow programs interpreter). Implementation objects are connected with each other via Qt signals and slots in the same order as corresponding blocks are connected on a diagram (<<Publisher-subscriber>> design pattern is used here). On the second step the interpreter executes each block that does not have any incoming data flows, and then each such block in its turn activates blocks that it has outgoing data flows to when the data tokens are ready. Blocks execution is performed in a pseudo-parallel way over a centralized message queue (see Fig.~\ref{image:dsPseudoparallelism}). This implementation detail distinguishes this language from all other industrial solutions (e.g. in Microsoft Robotics Developer Studio diagrams are deployed as a set of independent web services~\cite{jackson2007microsoft}). We choose this approach because this programming environment targets low-performance controllers of educational robots, that don't embrace true parallel execution of multiple programs at once. Nevertheless, the language has a process fork block that is useful while implementing complex control models mentioned above. This block should be used as a low-level tool for execution control.

@@

2D robot simulation environment is the most stand-alone TRIK Studio subsystem. Its window is a part of TRIK Studio's user interface, but it also can be used separately. The main part of this environment is an editor for virtual world model. Using a special menu one can select one of supported drawing tools (similar to how it is done in most graphical editors). There are two main drawing tools: drawing walls (solid objects that robots cannot move through and that are deteced by robots' ultrasound and infrared sensors) and colored markers on the floor (elements that are detected by color, light sensors and virtual robots' cameras). Colored markers could be drawn in a shape of lines, cubic Bézier curves, rectangles, ellipses or any shape drawn using mouse pointer. For each marker users are able to set their width, color and fill type.

The robot model itself is a differential two-wheeled truck. Robot model is always shown in this editor, users can place sensors onto it where they prefer. The regions where sensors scan for objects are highlighted. A separate panel is showing an image of the chosen controller (Lego NXT, EV3 or TRIK), which users can interact with: clicking on an image of a button emulates controller's button click event, users can manipulate LEDs placed on controller's top or output some text or images onto the controller's screen, for example, drawing shapes on it using drawing blocks (see Section~\ref{chapter:controlFlowLanguage}).

Running the program in the simulation environment is performed in the same editor where users created their models. World model can be changed even during program execution, putting additional walls or colored markers will take immediate effect of virtual robot's sensors. Positions and directions of the sensors themselves can also be changed in any time, which conforms with changing real robot's configuration while it is moving.

The simulation environment is built based on Model-View-Controller design pattern. The Model contains serialized specification of the robot and the world, and it notifies all other components if any property of any object is changed. The View is subscribed to these events, all updates are immediately displayed in the editor or other simulator windows. All user actions are handled by the Controller component, which uses an intermediate command stack to be able to undo such commands on demand.

The simulation environment has two implementations of robot behavior models: <<ideal>> and <<realistic>>. Within the ideal model all friction forces (for floor and walls) are ignored, and having constant motor powers the robot continues to move without any acceleration or slowdown. Any, even slightest collision with a wall will stop the robot. The realistic model takes care of thrust and friction force, when crashing into a wall the robot tries to behaves like the real device. Users is also have an ability to add Gaussian noise to sensor data and motor impulses. Changing between these models is done using a checkbox in simulation settings window and could be done even during the execution of a program.

Time in the simulation environment does not equal computer's time: there is a special centralized timeline, the speed of which can be controlled from a special panel. The same timeline is used in the diagram interpreters to match the work of timer blocks with virtual time independently of CPU's time.
